Learning words by recursive pragmatic reasoning about other agents

### 2013-05-31 praglearn.tex
more text tinkering, especially results

\begin{document}

\maketitle

@@

The goal of our current work is to investigate the possibilities for
integrating models of recursive pragmatic reasoning with models of
language learning, with the hope of capturing phenomena in both
domains. We begin by describing previous models in each domain as well
as proposals for bringing the two together. We next simulate findings
from the domains of pragmatic inference, word learning, and their
intersection. We end by discussing the implications of our results for
work on language change.

@@

\subsection{Paradoxes in optimal models of pragmatic learning}
% When modelling human behaviour, it's useful to begin with optimal
% models; this has the advantage that when we are then forced to
% deviate from the purely rational approach, it's generally in a
% theoretically interpretable way. However,
Multi-agent interactions are difficult to model in a normative or
optimal framework without falling prey to paradox. For example,
consider a simple model of the agents in the above game. First we
define a \textit{literal listener} $L_0$. This agent has a
\textit{lexicon} which probabilistically maps words to meanings;
specifically, it assigns each word $w$ a vector describing the extent
to which this word provides evidence for each object\footnote{We
  assume words refer directly to objects, rather than to abstract
  semantic features. Our simplification is without loss of
  generalization, however, because we can interpret our model as
  marginalizing over such a representation, with our literal
  $P_\lex(\obj | \word) = \sum_\text{features} P(\obj |
  \text{features}) P_\lex(\text{features} | \word)$.}. Each entry in
the vector is in $(0, 1)$, and together they sum to 1:
\begin{align*}
P_{L_0}(\obj | \word, \lex) &\propto \lex(\word, \obj) \times \prior.
\end{align*}
Confronted with such a listener, a probabilistically rational speaker
should attempt to choose a word which soft-maximizes the probability
that the listener will assign to the target object---modulated by the
effort or cost associated with producing this word:
\begin{align*}
P_{S_1}(\word | \obj, \lex) &\propto \exp\Big(\lambda \big(\log P_{L_0}(\obj | \word, \lex) - \text{cost}(\word)\big)\Big).
\end{align*}
But given this speaker, then the naive $L_0$ strategy is not
optimal. Instead, listeners should use Bayes rule to invert the
speaker's decision procedure \cite{frank2012}:
\begin{align*}
P_{L_2}(\obj | \word, \lex) &\propto P_{S_1}(\word | \obj, \lex) \times \prior.
\end{align*}
Now the paradox becomes apparent. Given such a listener, it is no
longer optimal for speakers to implement strategy $S_1$; instead, they
should implement strategy $S_3$ which soft-maximizes $P_{L_2}$ instead
of $P_{L_0}$. And then listeners ought to implement $L_4$, and so on.

One computationally expensive option is to continue iterating such
strategies until reaching a fixed point equilibrium. While this
strategy guarantees that each agent will behave normatively given the
other agent's strategy, there is no guarantee that such strategies
will be near the system's global optimum; equilibrium strategies make
up only a small subspace of all possible strategies. Nor is there much
evidence the humans find such equilibria; their behaviour in language
games (and in other games \cite{camerer2004}) can be well-modeled as
implementing $S_n$ or $L_n$ for some small $n$
\cite{frank2012}. Following this work, we recurse a finite (small)
number of times. The consequence is that one agent, implementing
$S_n$, is fully optimal with respect to the game, while the other,
implementing $L_{n-1}$, is nearly optimal---off by only a single
recursion.

As soon as we attempt to add uncertainty about the meanings of words
to such a model, new paradoxes arise. Suppose the listener is a young
child who is uncertain about the lexicon their partner is using. The
obvious solution is for them to place a prior on the lexicon; they
then update their posterior based on whatever utterances and
contextual cues they observe, and in the mean time interpret each
utterance by making their best guess, marginalizing out this
uncertainty. This basic structure is captured in previous models of
Bayesian word learning \cite{frank2009}. But when combined with the
recursive pragmatic model, a new question arises: Given such a
listener, what model should the speaker use? A rational speaker
attempts to maximize the listener's likelihood of understanding, so if
an uncertain listener interpets by marginalizing over some posterior,
then a knowledgeable speaker should calculate this posterior and
marginalize themselves. But then the speaker is not using any
particular lexicon at all, and a rational learning listener should
therefore not be trying to infer the speaker's lexicon.\footnote{Of
  course, in reality both parties will generally have some
  uncertainty, making the situation even worse. If we start from an
  uncertain listener with a prior over lexicons, then a first-level
  uncertain speaker needs a prior over priors on lexicons, a
  second-level uncertain listener needs a prior over priors over
  priors, etc. The original $L_0 \rightarrow S_1 \rightarrow \dots$
  recursion was bad enough, but at least each step had a constant
  cost. This new recursion produces hyper$^n$-distributions for which
  inference almost immediately becomes intractable even in principle,
  since the dimensionality of the learning problem increases with each
  step. Yet, without this addition of new uncertainty at each level,
  the model would dissolve back into certainty as in the previous
  paragraph, making learning
  impossible.}
% In fact, such an uncertain listener's uncertainty dissolves; they
% may not know the lexicon, but they do
% know their posterior over the lexicon -- and because
% this is all that the speaker cares about, there is
% nothing for the listener to learn from the speaker's
% behaviour.

@@

Assumption (a) corresponds to the lexicographer's illusion: Naive
language users will argue vociferously that words have specific
meanings, even though these meanings are unobservable to everyone who
purportedly uses them. It also explains why learners speak the
language they hear (rather than some private language that they assume
listeners will eventually learn): Under assumption (a), observing
other speakers' behaviour provides data about not just that speaker's
idiosyncratic lexicon, but the consensus lexicon. Assumption (b)
avoids the explosion of hyper$^n$-distributions described above: If
agent $n$ knows the lexicon, they assume that all lower agents do as
well, reducing to the original tractable model without
uncertainty. And assumption (c) introduces a limited form of
uncertainty, and thus the potential for learning.
% Everything else follows by applying standard Bayesian reasoning
% given these assumptions.

@@

In several important cases, these assumptions are as close to rational
as possible within the finite-recursion framework. If $L_n(\obj |
\word, \text{$S$'s data})$ and $L_n(\obj | \word, \text{$L$'s data})$
coincide, then $S$ is fully normative and $L$ is off by only a single
recursion. This case in particular includes all previous experiments
using this pragmatic reasoning framework, which considered only
single-turn interactions where $\text{$L$'s data} = \text{$S$'s data}
= \emptyset$. Alternatively, if one agent has enough data that their
uncertainty is effectively non-existent, then it becomes normative for
their partner to treat them as being fully knowledgeable, and this is
a reasonable description of the child language learning case. In
general, though, in arbitrary multi-turn interactions in which both
agents have non-trivial uncertainty, this model induces complex and
non-normative learning dynamics. In the remainder of this paper, we
investigate the model's behaviour in all three situations.\footnote{In
  our simulations throughout, we somewhat arbitrarily set $n = 2$ (the
  minimal value that produces all the qualitative phenomena), $\lambda
  = 3$, and assume that all agents have shared priors on the lexicon
  and full knowledge of the cost function. Inference is via importance
  sampling.}

@@

\section{Results: Pure pragmatic inference}

@@

We begin by applying the model described above to a set of one-shot
pragmatic inference games that have been well-studied in linguistics
\cite{grice1975,horn1984} and are addressed by previous one-shot
models of pragmatic inference \cite{frank2012,bergen2012}. These
situations set the stage for simulations investigating how learning
proceeds in iterated versions of such games, described in the
following section. Results captured by our model and previous models
are summarized in Table \ref{tab:results}.

@@

% XX note the relation to BGL model; they need uncertainty in the
% middle of the pragmatics specifically, which is where we end up
% putting it for other reasons. And in Bayesian framework uncertainty
% and learning must always coincide. Also the intuition that this is a
% general property of uncertainty for soft-max agents -- if you don't
% know where you are, good worlds have more peaked utility
% distributions than bad worlds, and thus more consistent behaviour by
% other agents, so on average when you're guessing what other agents
% will do, good worlds contribute more to your average than bad
% worlds.

@@

Humans, however, unlike traditional game theoretic models, do make the
inference that if two utterances seem equivalent except that one is
more costly to utter, then the costly utterance refers to an unusual
event. We call this pattern \textit{Horn implicature}, after
\cite{horn1984}. For instance, ``Lee got the car to stop'' implies
that Lee used an unusual method (e.g.~not the brakes) because, had he
used the brakes, the speaker would have chosen the simpler and shorter
(less costly) expression, ``Lee stopped the car''
\cite{horn1984}. Surprisingly, Bergen et al \cite{bergen2012} show
that the key to achieving this favorable result is ignorance. If a
listener assigns equal probability to her partner using the good
system or the bad system, then their best bet is to estimate
$P_S(\word | \obj)$ as the average of $P_S(\word | \obj, \text{good
  system})$ and $P_S(\word | \obj, \text{bad system})$. These might
seem to cancel out, but in fact they do not. In the good system, the
speaker has higher expected utility than the bad system; therefore, a
soft-max agent in the bad system has noiser behaviour than in the good
system, and their behaviour in the good system dominates the average,
breaking the symmetry. Similar reasoning applies to an uncertain
speaker.

@@

The model of \cite{bergen2012} differs from ours in that they then
apply further layers of pragmatic recursion on top of these marginal
distributions, which as described above removes the possibility of
learning; in their model it is only hypothetical agents who are
ignorant, while the actual speaker and listener know everything there
is to know about each other's behaviour. However, the core result
carries over to our model. For example, with a uniform prior over
lexicons and $P_\text{prior}(\text{\textsc{common}}) = 0.8$,
$\text{cost}(\text{``cheap''}) = 0.5$,
$\text{cost}(\text{``expensive''}) = 1.0$, the symmetry breaks in the
appropriate way: Despite total ignorance about the conventional
system, speakers prefer to use simple words for common referents
($P_S(\text{``cheap''} | \text{\textsc{common}}) = 0.88,
P_S(\text{``cheap''} | \text{\textsc{rare}}) = 0.46$), and listeners
show a similar bias ($P_L(\text{\textsc{common}} | \text{``cheap''}) =
0.77, P_L(\text{\textsc{common}} | \text{``expensive''}) =
0.65$). This preference is somewhat weak; the critical observation is
that it exists at all, given the unbiased priors. We return to this
point in \S\ref{sec:horn-emergence}.

@@

\section{Results: The role of pragmatics in language acquisition}

@@

\begin{SCfigure}[1]
  \centering
  % \includegraphics[width=0.24\textwidth]{figures/ME-1dax.pdf}
  % \includegraphics[width=0.24\textwidth]{figures/ME-flat-10dog-10dax.pdf}
  % \includegraphics[width=0.24\textwidth]{figures/ME-antisparse-10dog-10dax.pdf}
  % \includegraphics[width=0.24\textwidth]{figures/ME-sparse-10dog-10dax.pdf}
\includegraphics[width=.6\textwidth]{figures/ME-composite-small.pdf}
  \caption{Results for simulations of the ``disambiguation'' phenomenon. A novel word is used in a situation with a novel object and a familiar object. Each plot shows the joint posterior probability that the familiar word refers to the familiar object (horizontal) and that the novel word refers to the familiar object (vertical). Left plot shows the posterior after a single situation; right plot shows multiple situations.
%The right middle plot shows the posterior after a single situation given either a prior that makes it likely that there are words referring to both the novel and familiar objects (perhaps because the two objects are physically similar; the right plot shows a situation where the prior makes it unlikely that a word refers to both (objects dissimilar). 
  }
  \label{fig:me}
\end{SCfigure}

Children, when presented with both a novel and a familiar object
(e.g. an eggbeater and a ball), will treat a novel label
(e.g. ``dax'') as referring to the novel object, for example by
supplying the eggbeater when asked to ``give me the dax''
\cite{markman1988}. This phenomenon is sometimes referred to as
``mutual exclusivity'' or even ``Fast mapping,'' and many distinct
theoretical accounts have been proposed in the psychological
literature. Simple probabilistic word learning models can produce a
similar pattern of findings \cite{frank2009}, often due to the kinds
of ``explaining away'' type phenomena that are observed in such
models. But all such models assume that learners in fact retain the
mapping between word and object that is demonstrated in the
experimental situation; this observation is contradicted, however, by
evidence that children often do not retain the mappings that are
demonstrated by their inferences in the moment \cite{horst2008}.

Our model provides an intriguing possible explanation of this finding:
given a single disambiguation situation, the model gives a substantial
probability (e.g. 75\%) that the speaker is referring to the novel
object. Nevertheless, this is not accompanied by an increased belief
that the novel word literally refers to this object (as shown in
Fig. \ref{fig:me}, left). This comes in part from a potential
specificity implicature: the listener knows that the familiar word
\emph{does not} refer to the novel object--hence the novel word (even
if it refers to the familiar object as well) is the best way to refer
to the novel object. Nevertheless, on repeated exposure to the same
novel word, novel object situation, the model does lexicalize this
inference (Fig. \ref{fig:me}, right).

@@

\begin{SCfigure}[1]
\centering
\includegraphics[width=1.5in]{figures/some-all-only-pragmatic.pdf}
\includegraphics[width=1.5in]{figures/some-all-pragmatic+unambiguous.pdf}
\caption{\label{fig:scalar} Simulations..}
\end{SCfigure}

@@

The acquisition of scalar adjectives like ``some'' provides a puzzle
for most models of word learning: given that in many contexts, the
word ``some'' is used to mean \textsc{some but not all}, how do
children learn that this is not in fact its literal meaning? Our model
is able to take specificity implicatures into account when learning,
and thus potentially answer this puzzle. Imagine two bowls: one
containing a mixture of apples and oranges, the other containing
apples alone. Fig.~\ref{fig:scalar} (left) shows the model's beliefs
after encountering the words ``some'' and ``all'' used contrastively
to refer to these two bowls (as in e.g.\ ``the bowl with some
apples''). It believes that at least one of these words must have a
restrictive literal meaning, but not both -- it leaves open the
possibility that either ``some'' or ``all'' might be literally
ambiguous. Fig.~\ref{fig:scalar} (right) then shows the model's
beliefs if we also provide examples of ``some'' referring to exactly
the same object, but in contexts where this is the only object
available, so there is no contrast and thus no implicature; this
convinces the model that ``all apple'' literally refers only to the
bowl that contains no oranges, while ``some'' most likely is somewhat
ambiguous. Crucially, the model is able to learn this without once
ever observing ``some'' used

\section{Results: Mutual learning}

@@

Experimental results suggest that communicators are able to establish
novel, consensus-based communication systems. For example, adults
playing a communication game using only novel symbols with no
conventional meaning will typically converge on a set of new
conventions which allow them to accomplish their task
\cite{galantucci2005}. Or in a less extreme example, communicators
asked to refer to novel objects invent conventional names for them
over the course of repeated interactions (e.g., ``the ice skater'' for
an abstract figure vaguely resembling an ice skater,
\cite{clark1986}). From a pure learning perspective this is anomalous,
however: Since both agents know perfectly well that there is no
underlying convention to discover, there is nothing to learn from the
other's behaviour. Furthermore, even if only one partner is producing
the novel expressions, their behaviour still becomes more regular
(conventional) over time, which would seem to rule out a role for
learning---even if there is some pattern in the expressions the
speaker chooses to use, there is certainly nothing for the
\textit{speaker} to learn by observing these patterns, and thus their
behaviour should not change over time.
% An obvious hack might be to simply stipulate that speakers take
% their own utterances as training data for further learning, but this
% would fail to explain yet another wrinkle in this phenomenon:
% speakers only regularize their production over time if they have
% some kind of feedback indicating that the listener understands; if
% such feedback is removed, the phenomenon is dramatically reduced (XX
% krauss 1966, branigan catchpole pickering 2011).

@@

Pragmatic inference here does two things. First, the pragmatic
recursion makes it possible for the speaker to learn from the
listener's interpretations, which would be impossible without an
explicit model of the listener as a rational agent. Second, pragmatics
creates the sparse structure in the observed lexicons: At each step,
the goal-directed speaker attempts to exploit whatever communicatively
useful asymmetries exist in its current model of the listener, and
these attempts tend to be successful. If the listener has even a small
asymmetry in their current posterior then their pragmatic reasoning
will magnify it as well.

@@

\begin{SCfigure}
\centering
\includegraphics[width=0.5\textwidth]{figures/horn-composite.pdf}
% \includegraphics[width=0.48\textwidth]{figures/horn-emergence-1.pdf}
\caption{\label{fig:horn} Example simulations showing the lexicalization of horn implicatures. Plotting conventions are as above.}
\end{SCfigure}

>>>>

\begin{document}
\maketitle

@@

The goal of our current work is to investigate the possibilities for
integrating models of recursive pragmatic reasoning with models of
language learning, with the hope of capturing phenomena in both
domains. We begin by describing previous models in each domain as well
as proposals for bringing the two together. We next simulate findings
on pragmatic inference in one-shot games (replicating previous
work). We then build on these results to simulate the results of pragmatic
learning in the language acquisition setting where one communicator is
uncertain about the lexicon and in iterated communication games where both
communicators are uncertain about the lexicon. We end by discussing the implications of our results for
work on language change.

@@

\subsection{Paradoxes in optimal models of pragmatic learning}
% When modelling human behavior, it's useful to begin with optimal
% models; this has the advantage that when we are then forced to
% deviate from the purely rational approach, it's generally in a
% theoretically interpretable way. However,
Multi-agent interactions are difficult to model in a normative or
optimal framework without falling prey to paradox. For example,
consider a simple model of the agents in the above game. First we
define a \textit{literal listener} $L_0$. This agent has a
\textit{lexicon} which probabilistically maps words to meanings;
specifically, it assigns each word $w$ a vector describing the extent
to which this word provides evidence for each object\footnote{We
  assume words refer directly to objects, rather than to abstract
  semantic features. Our simplification is without loss of
  generalization, however, because we can interpret our model as
  marginalizing over such a representation, with our literal
  $P_\lex(\obj | \word) = \sum_\text{features} P(\obj |
  \text{features}) P_\lex(\text{features} | \word)$.}. Each entry in
the vector is in $(0, 1)$, and together they sum to 1:
\begin{align*}
P_{L_0}(\obj | \word, \lex) &\propto \lex(\word, \obj) \times \prior.
\end{align*}
Confronted with such a listener, a probabilistically rational speaker
should attempt to choose a word which soft-maximizes the probability
that the listener will assign to the target object---modulated by the
effort or cost associated with producing this word:
\begin{align*}
P_{S_1}(\word | \obj, \lex) &\propto \exp\Big(\lambda \big(\log P_{L_0}(\obj | \word, \lex) - \text{cost}(\word)\big)\Big).
\end{align*}
But given this speaker, then the naive $L_0$ strategy is not
optimal. Instead, listeners should use Bayes rule to invert the
speaker's decision procedure \cite{frank2012}:
\begin{align*}
P_{L_2}(\obj | \word, \lex) &\propto P_{S_1}(\word | \obj, \lex) \times \prior.
\end{align*}
Now the paradox becomes apparent. Given such a listener, it is no
longer optimal for speakers to implement strategy $S_1$; instead, they
should implement strategy $S_3$ which soft-maximizes $P_{L_2}$ instead
of $P_{L_0}$. And then listeners ought to implement $L_4$, and so on.

One computationally expensive option is to continue iterating such
strategies until reaching a fixed point equilibrium. While this
strategy guarantees that each agent will behave normatively given the
other agent's strategy, there is no guarantee that such strategies
will be near the system's global optimum; equilibrium strategies make
up only a small subspace of all possible strategies. Nor is there much
evidence the humans find such equilibria; their behavior in language
games (and in other games \cite{camerer2004}) can be well-modeled as
implementing $S_n$ or $L_n$ for some small $n$
\cite{frank2012}. Following this work, we recurse a finite (small)
number of times. The consequence is that one agent, implementing
$S_n$, is fully optimal with respect to the game, while the other,
implementing $L_{n-1}$, is nearly optimal---off by only a single
recursion.

As soon as we attempt to add uncertainty about the meanings of words
to such a model, new paradoxes arise. Suppose the listener is a young
child who is uncertain about the lexicon their partner is using. The
obvious solution is for them to place a prior on the lexicon; they
then update their posterior based on whatever utterances and
contextual cues they observe, and in the mean time interpret each
utterance by making their best guess, marginalizing out this
uncertainty. This basic structure is captured in previous models of
Bayesian word learning \cite{frank2009}. But when combined with the
recursive pragmatic model, a new question arises: Given such a
listener, what model should the speaker use? A rational speaker
attempts to maximize the listener's likelihood of understanding, so if
an uncertain listener interpets by marginalizing over some posterior,
then a knowledgeable speaker should calculate this posterior and
marginalize themselves. But then the speaker is not using any
particular lexicon at all, and a rational learning listener should
therefore not be trying to infer the speaker's lexicon.\footnote{Of
  course, in reality both parties will generally have some
  uncertainty, making the situation even worse. If we start from an
  uncertain listener with a prior over lexicons, then a first-level
  uncertain speaker needs a prior over priors on lexicons, a
  second-level uncertain listener needs a prior over priors over
  priors, etc. The original $L_0 \rightarrow S_1 \rightarrow \dots$
  recursion was bad enough, but at least each step had a constant
  cost. This new recursion produces hyper$^n$-distributions for which
  inference almost immediately becomes intractable even in principle,
  since the dimensionality of the learning problem increases with each
  step. Yet, without this addition of new uncertainty at each level,
  the model would dissolve back into certainty as in the previous
  paragraph, making learning
  impossible.}
% In fact, such an uncertain listener's uncertainty dissolves; they
% may not know the lexicon, but they do
% know their posterior over the lexicon -- and because
% this is all that the speaker cares about, there is
% nothing for the listener to learn from the speaker's
% behavior.

@@

Assumption (a) corresponds to the lexicographer's illusion: Naive
language users will argue vociferously that words have specific
meanings, even though these meanings are unobservable to everyone who
purportedly uses them. It also explains why learners speak the
language they hear (rather than some private language that they assume
listeners will eventually learn): Under assumption (a), observing
other speakers' behavior provides data about not just that speaker's
idiosyncratic lexicon, but the consensus lexicon. Assumption (b)
avoids the explosion of hyper$^n$-distributions described above: If
agent $n$ knows the lexicon, they assume that all lower agents do as
well, reducing to the original tractable model without
uncertainty. And assumption (c) introduces a limited form of
uncertainty, and thus the potential for learning.
% Everything else follows by applying standard Bayesian reasoning
% given these assumptions.

@@

In several important cases, these assumptions are as close to rational
as possible within the finite-recursion framework. If $L_n(\obj |
\word, \text{$S$'s data})$ and $L_n(\obj | \word, \text{$L$'s data})$
coincide, then $S$ is fully normative and $L$ is off by only a single
recursion. This case in particular includes all previous experiments
using this pragmatic reasoning framework, which considered only
single-turn interactions where $\text{$L$'s data} = \text{$S$'s data}
= \emptyset$. Alternatively, if one agent has enough data that their
uncertainty is effectively non-existent, then it becomes normative for
their partner to treat them as being fully knowledgeable, and this is
a reasonable description of the child language learning case. In
general, though, in arbitrary multi-turn interactions in which both
agents have non-trivial uncertainty, this model induces complex and
non-normative learning dynamics.

In the remainder of the paper, we apply the model described above to a set of one-shot
pragmatic inference games that have been well-studied in linguistics
\cite{grice1975,horn1984} and are addressed by previous one-shot
models of pragmatic inference \cite{frank2012,bergen2012}. These
situations set the stage for simulations investigating how learning
proceeds in iterated versions of such games, described in the
following section. Results captured by our model and previous models
are summarized in Table \ref{tab:results}. In
  our simulations throughout, we somewhat arbitrarily set $n = 2$ (the
  minimal value that produces all the qualitative phenomena), $\lambda
  = 3$, and assume that all agents have shared priors on the lexicon
  and full knowledge of the cost function. Inference is via importance
  sampling.

@@

\section{Pragmatic inference in one-shot games}

@@

@@

% XX note the relation to BGL model; they need uncertainty in the
% middle of the pragmatics specifically, which is where we end up
% putting it for other reasons. And in Bayesian framework uncertainty
% and learning must always coincide. Also the intuition that this is a
% general property of uncertainty for soft-max agents -- if you don't
% know where you are, good worlds have more peaked utility
% distributions than bad worlds, and thus more consistent behavior by
% other agents, so on average when you're guessing what other agents
% will do, good worlds contribute more to your average than bad
% worlds.

@@

Humans, however, unlike traditional game theoretic models, do make the
inference that if two utterances seem equivalent except that one is
more costly to utter, then the costly utterance refers to an unusual
event. We call this pattern \textit{Horn implicature}, after
\cite{horn1984}. For instance, ``Lee got the car to stop'' implies
that Lee used an unusual method (e.g.~not the brakes) because, had he
used the brakes, the speaker would have chosen the simpler and shorter
(less costly) expression, ``Lee stopped the car''
\cite{horn1984}. Surprisingly, Bergen et al \cite{bergen2012} show
that the key to achieving this favorable result is ignorance. If a
listener assigns equal probability to her partner using the good
system or the bad system, then their best bet is to estimate
$P_S(\word | \obj)$ as the average of $P_S(\word | \obj, \text{good
  system})$ and $P_S(\word | \obj, \text{bad system})$. These might
seem to cancel out, but in fact they do not. In the good system, the
speaker has higher expected utility than the bad system; therefore, a
soft-max agent in the bad system has noiser behavior than in the good
system, and their behavior in the good system dominates the average,
breaking the symmetry. Similar reasoning applies to an uncertain
speaker.

@@

The model of \cite{bergen2012} differs from ours in that they then
apply further layers of pragmatic recursion on top of these marginal
distributions, which as described above removes the possibility of
learning; in their model it is only hypothetical agents who are
ignorant, while the actual speaker and listener know everything there
is to know about each other's behavior. However, the core result
carries over to our model. For example, with a uniform prior over
lexicons and $P_\text{prior}(\text{\textsc{common}}) = 0.8$,
$\text{cost}(\text{``cheap''}) = 0.5$,
$\text{cost}(\text{``expensive''}) = 1.0$, the symmetry breaks in the
appropriate way: Despite total ignorance about the conventional
system, speakers prefer to use simple words for common referents
($P_S(\text{``cheap''} | \text{\textsc{common}}) = 0.88,
P_S(\text{``cheap''} | \text{\textsc{rare}}) = 0.46$), and listeners
show a similar bias ($P_L(\text{\textsc{common}} | \text{``cheap''}) =
0.77, P_L(\text{\textsc{common}} | \text{``expensive''}) =
0.65$). This preference is somewhat weak; the critical observation is
that it exists at all, given the unbiased priors. We return to this
point in \S\ref{sec:horn-emergence}.

@@

\section{The role of pragmatics in language acquisition}

@@

% \begin{SCfigure}[1]
%   \centering
%   % \includegraphics[width=0.24\textwidth]{figures/ME-1dax.pdf}
%   % \includegraphics[width=0.24\textwidth]{figures/ME-flat-10dog-10dax.pdf}
%   % \includegraphics[width=0.24\textwidth]{figures/ME-antisparse-10dog-10dax.pdf}
%   % \includegraphics[width=0.24\textwidth]{figures/ME-sparse-10dog-10dax.pdf}
% \includegraphics[width=.6\textwidth]{figures/ME-composite-small.pdf}
%   \caption{Results for simulations of the ``disambiguation'' phenomenon. A novel word is used in a situation with a novel object and a familiar object. Each plot shows the joint posterior probability that the familiar word refers to the familiar object (horizontal) and that the novel word refers to the familiar object (vertical). Left plot shows the posterior after a single situation; right plot shows multiple situations.
% %The right middle plot shows the posterior after a single situation given either a prior that makes it likely that there are words referring to both the novel and familiar objects (perhaps because the two objects are physically similar; the right plot shows a situation where the prior makes it unlikely that a word refers to both (objects dissimilar). 
%   }
%   \label{fig:me}
% \end{SCfigure}

Children, when presented with both a novel and a familiar object
(e.g. an eggbeater and a ball), will treat a novel label
(e.g. ``dax'') as referring to the novel object, for example by
supplying the eggbeater when asked to ``give me the dax''
\cite{markman1988}. This phenomenon is sometimes referred to as
``mutual exclusivity'' or even ``fast mapping.'' Simple probabilistic word learning models can produce a
similar pattern of findings \cite{frank2009}, but all such models assume that learners retain the
mapping between novel word and novel object demonstrated in the
experimental situation. This observation is contradicted, however, by
evidence that children often do not retain the mappings that are
demonstrated by their inferences in the moment \cite{horst2008}.

Our model provides an intriguing possible explanation of this finding:
given a single disambiguation situation, the model gives a substantial
probability (e.g. 75\%) that the speaker is referring to the novel
object. Nevertheless, this inference is not accompanied by an increased belief
that the novel word literally refers to this object. The learner's
interpretation arises not from lexical mapping but instead from a kind
of 
specificity implicature: the listener knows that the familiar word
\emph{does not} refer to the novel object---hence the novel word (even
if it refers to the familiar object as well) is the best way to refer
to the novel object. Nevertheless, on repeated exposure to the same
novel word, novel object situation, the learner does learn the mapping
as part of the lexicon (congruent with other data on repeated training
on disambiguation situations \cite{bion2012}).

@@

% \begin{SCfigure}[1]
% \centering
% \includegraphics[width=1.5in]{figures/some-all-only-pragmatic.pdf}
% \includegraphics[width=1.5in]{figures/some-all-pragmatic+unambiguous.pdf}
% \caption{\label{fig:scalar} Simulations..}
% \end{SCfigure}

@@

The acquisition of scalar adjectives like ``some'' provides a puzzle
for most models of word learning: given that in many contexts, the
word ``some'' is used to mean \textsc{some but not all}, how do
children learn that {\sc some but not all} is not in fact its literal meaning? Our model
is able to take specificity implicatures into account when learning,
and thus provide a potential solution.

Imagine two bowls: one
containing a mixture of apples and oranges, the other containing
apples alone. After encountering the words ``some'' and ``all'' used contrastively
to refer to these two bowls (as in e.g.\ ``the bowl with some
apples''), a learner under our model believes that at least one of these words must have a
restrictive literal meaning, but not both---it leaves open the
possibility that either ``some'' or ``all'' might be literally
ambiguous. But if examples of ``some'' are given that refer to exactly
the same object in contexts where this is the only object
available, so there is no contrast and thus no implicature; this
convinces the model that ``all apple'' literally refers only to the
bowl that contains no oranges, while ``some'' most likely is somewhat
ambiguous. Crucially, the model is able to learn this without once
ever observing ``some'' used

\section{Mutual learning situations}

@@

Experimental results suggest that communicators are able to establish
novel, consensus-based communication systems. For example, adults
playing a communication game using only novel symbols with no
conventional meaning will typically converge on a set of new
conventions which allow them to accomplish their task
\cite{galantucci2005}. Or in a less extreme example, communicators
asked to refer to novel objects invent conventional names for them
over the course of repeated interactions (e.g., ``the ice skater'' for
an abstract figure vaguely resembling an ice skater,
\cite{clark1986}). From a pure learning perspective this behavior is anomalous,
however: Since both agents know perfectly well that there is no
underlying convention to discover, there is nothing to learn from the
other's behavior. Furthermore, even if only one partner is producing
the novel expressions, their behavior still becomes more regular
(conventional) over time, which would seem to rule out a role for
learning---even if there is some pattern in the expressions the
speaker chooses to use, there is certainly nothing for the
\textit{speaker} to learn by observing these patterns, and thus their
behavior should not change over time.
% An obvious hack might be to simply stipulate that speakers take
% their own utterances as training data for further learning, but this
% would fail to explain yet another wrinkle in this phenomenon:
% speakers only regularize their production over time if they have
% some kind of feedback indicating that the listener understands; if
% such feedback is removed, the phenomenon is dramatically reduced (XX
% krauss 1966, branigan catchpole pickering 2011).

@@

% Pragmatic inference here does two things. First, the pragmatic
% recursion makes it possible for the speaker to learn from the
% listener's interpretations, which would be impossible without an
% explicit model of the listener as a rational agent. Second, pragmatics
% creates the sparse structure in the observed lexicons: At each step,
% the goal-directed speaker attempts to exploit whatever communicatively
% useful asymmetries exist in its current model of the listener, and
% these attempts tend to be successful. If the listener has even a small
% asymmetry in their current posterior then their pragmatic reasoning
% will magnify it as well.

@@

\begin{SCfigure}
\centering
\includegraphics[width=0.5\textwidth]{figures/horn-composite.pdf}
% \includegraphics[width=0.48\textwidth]{figures/horn-emergence-1.pdf}
\caption{\label{fig:horn} Example simulations showing the
  lexicalization of Horn implicatures. Plotting conventions are as
  above. In the first run, speaker and listener converge on a sparse
  and  efficient communicative equilibrium, while in the second they reach
  a sub-optimal equilibrium.}
\end{SCfigure}

\begin{SCfigure}
\centering
\includegraphics[width=0.3\textwidth]{figures/emergence-average.pdf}
\hspace{0.2in}
\includegraphics[width=0.3\textwidth]{figures/emergence-horn-average.pdf}
\caption{\label{fig:emergence-average} Averaged behavior over 300
  dialogues as in Figs.~\ref{fig:emergence} and \ref{fig:horn}. Left:
  Communicative success by game type and dialogue turn. Right:
  Proportion of dyads in the Horn implicature game
  (\S\ref{sec:horn-emergence}) who have converged on the ``good'' or
  ``bad'' lexicons and believe that these are literal meanings.}
\end{SCfigure}

A stronger example of how pragmatics can create communicative biases can be
observed by considering the Horn implicatures introduced in \S\ref{sec:horn-implicature}. 
In these games, pragmatic reasoning causes each agent to start out weakly biasedtowards the associations ``cheap'' $\leftrightarrow$
\textsc{common}, ``expensive'' $\leftrightarrow$ \textsc{rare}; this bias arises because their prior on literal meaning
is uniform and uninformative. A fully rational listener, therefore,
observing an uncertain speaker using this mapping, would discount it
as arising from this bias, and conclude that the speaker in fact did
not know which convention was in use. Conventional listeners,
however, believe that speakers do know which convention is in
use, and therefore tend to take their behavior as positive evidence that the
``good'' system is in use. Similarly, conventional speakers
will wager that since on average they will succed more often if
listeners are using the ``good'' system, they might as well try
it. When they succeed, they take their success as evidence that the listener was in
fact using the good system all along. As a result, dyads in this game
not only end up converging onto a stable system at a rate far above
chance (Fig.~\ref{fig:horn}.

Something
fundamentally different happens in this case than in the specificity
implicature case (\S\ref{sec:learning-specificity-implic}), however. There, we
saw that our model could hear pragmatically strengthened usages
of words while recovering the underlying literal meaning. In the Horn
case, though, the pragmatic strengthening depends essentially on
ignorance about literal meaning. As the agents gather more data,
the implicature disappears, to be replaced by a belief that
``cheap'' \textit{literally} means \textsc{common}. To demonstrate
this phenomenon, we queried each agent in each dyad about how they would refer to
or interpret each object and word, \textit{if} the two objects were
equally common, which cancels the Horn implicature. As shown in
Fig.~\ref{fig:emergence-average} (right), after 30 turns, in nearly 70\%
of dyads both $S$ and $L$ believed in the ``good'' mapping even in
this implicature-free case, while less than 20\% believed in the
``bad'' mapping (with the rest being inconsistent).
